Query Optimization for Dynamic Imputation

### 2017-02-28 paper/algorithm.tex
Update imputation strategy session
Clarify relation to multiple imputation

\ProjectName{} uses a general-purpose imputation strategy based on chained-equation classification and regression trees (CE-CART).
Chained equation imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting a predictive model of one missing attribute which is conditional on all complete attributes and the other missing attributes.
Decision tree algorithms like CART are found to be effective~\cite{akande2015empirical} and are widely used in epidemiological domains~\cite{burgette2010multiple}.

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal effort and without changes to the optimizer, simply by adding time and loss functions (\Cref{sec:cost}) which describe the new algorithm.
In principle, new imputation algorithms can be added which are targeted to a specific domain.

These imputation algorithms are rarely used in isolation --- rather, the ultimate goal is to
perform statistical analysis of the complete data. To this end, the technique of
\textit{multiple imputation} can be used, in which multiple distinct copies of the complete
data are generated and estimators are computed by averaging over the multiple datasets. In
ImputeDB, we cannot assume the user desires multiple copies of the query result, though we
note that in further work, multiple imputation could be used within ImputeDB to reduce the
variance of estimates of some aggregates, like averages.

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change across
epochs) or a fixed number of epochs is reached.

The flexibility of this non-parametric imputation approach aligns well with the spirit of
query optimization, which aims to reconcile performance with a declarative interface to data
manipulation.

It is worth noting that given the nature of the imputation strategy used here, imputation
operators become blocking in our system. An extension of ImputeDB could consider regression
algorithms that learn in an online nature; this would reduce the cost of the imputation at
the possible expense of quality. Again, given the imputation strategy-agnostic design of
ImputeDB, this would not pose a problem.

>>>>

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal
effort and without changes to the optimizer, simply by adding time and loss functions
(\Cref{sec:cost}) which describe the new algorithm. In principle, new imputation algorithms
can be added, possibly targeted to a specific domain. The flexibility of this approach
aligns well with the spirit of query optimization, which aims to reconcile performance with
a declarative interface to data manipulation.

As a reference implementation, \ProjectName{} uses a general-purpose imputation strategy
based on chained-equation decision trees.  Chained equation
imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting
a predictive model of one missing attribute, conditional on all complete attributes
and all other missing attributes. Chained-equation decision tree algorithms are found to be
effective~\cite{akande2015empirical} and are widely used in epidemiological
domains~\cite{burgette2010multiple}. For our experiments, we use Weka's {\tt REPTree}
implementation (\cite{witten2016data}).

Our imputation algorithm proceeds by iteratively fitting regression trees to a subset of the
data and the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each {\it epoch}, the quality of the imputation
improves as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change
across epochs) or a fixed number of epochs is reached.

These imputation algorithms are used as part of a larger \textit{multiple imputation}
process, in which multiple distinct copies of the complete data are generated and estimators
are computed by averaging over the multiple datasets to form a single completed dataset
(\cite{gelman2006data}). In \ProjectName{}, we consider only single imputation, though in
further work, multiple imputation could be considered as another avenue of assessing
imputation confidence.

Given the nature of the imputation strategy used here, {\it Impute} operators are blocking
in our system. An extension of \ProjectName{} could consider regression algorithms that learn in
an online nature; this would reduce the cost of the imputation at the possible expense of
quality.
